---
title: "PML: Human Motion Project"
author: "Ben Apple"
date: "Thursday, December 19, 2014"
output: html_document
---

---
title: "Practical Machine Learning: HUman Motion Project"
output: html_document
---
### Executive Summary
Using a Random Forest classification algorithm, it was possible to achieve an estimated 1.5% sample error rate with a Kappa value of .98 these vaues indicate a well fitting model. The original data set contained 160 variables for this model that number was redueced to 53 variables of interest. It is ecpected that the Out of Sample Error rate will be 99% with a 95% confidence interval.

When the model was applied against the test data set the model produced a 20 out of 20 correct predictions. 

### Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. The goal of this project is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. The participants were asked to perform barbell lifts correctly and incorrectly in 5 different ways. The data from accelerometer is used in the following study. 

### Project Description
Participants in a study were asked to perform various weight lifting exercises while wearing accelerometers (including accelerometers on the weights).  The data was used to try to predict if the exercises where done correctly.  A full description of the test can be found here: http://groupware.les.inf.puc-rio.br/har.  

For this project, we were provided two sets of data collected from accelerometers worn by participants (and on the weights) while they performed a barbell lift. They were instructed how to do it correctly, and also how to do it incorrectly in 5 different ways. 

Using this data, this project attempts to predict whether the exercise was done correctly, or incorrectly in one of 5 different ways. 

### Data Analysis and Grooming
Two sets of accelerometer readings data were provided. The first was a training set was 19,622 observations of 160 variables, including the predicted variable indicating correct or incorrect technique. The second set was 20 observations of the same 160 variables, with the variable indicating correct or incorrect technique replaced with a dummy variable. 

Upon observation of the Training data, it was noted that there were variables which were empty or had values of NA for all observations.  These variables were removed from the training data with the corresponding variables removed from the testing data. The first seven variables of both data sets contained data which was not relevant to the model therefore these variables were removed. The irrelevant variables were number, user name, raw time 1, raw time 2, converted time, and two variables relating to window information.

The remaining 53 variable in the respective data sets were deamed as relevant. Using the training data, the variables were tested for Near Zero Covariates. None were found.  No further transformation or manipulation of the data took place. Below are the rsults of the test.


```{r, eval=FALSE}
nsv <- nearZeroVar(training, saveMetrics=T)
nsv
```

```
                     freqRatio percentUnique zeroVar  
roll_belt             1.101904     6.7781062   FALSE  
pitch_belt            1.036082     9.3772296   FALSE  
yaw_belt              1.058480     9.9734991   FALSE  
total_accel_belt      1.063160     0.1477933   FALSE  
gyros_belt_x          1.058651     0.7134849   FALSE  
gyros_belt_y          1.144000     0.3516461   FALSE  
gyros_belt_z          1.066214     0.8612782   FALSE  
accel_belt_x          1.055412     0.8357966   FALSE  
accel_belt_y          1.113725     0.7287738   FALSE  
accel_belt_z          1.078767     1.5237998   FALSE  
magnet_belt_x         1.090141     1.6664968   FALSE  
magnet_belt_y         1.099688     1.5187035   FALSE  
magnet_belt_z         1.006369     2.3290184   FALSE  
roll_arm             52.338462    13.5256345   FALSE  
pitch_arm            87.256410    15.7323412   FALSE  
yaw_arm              33.029126    14.6570176   FALSE  
total_accel_arm       1.024526     0.3363572   FALSE  
gyros_arm_x           1.015504     3.2769341   FALSE  
gyros_arm_y           1.454369     1.9162165   FALSE  
gyros_arm_z           1.110687     1.2638875   FALSE  
accel_arm_x           1.017341     3.9598410   FALSE  
accel_arm_y           1.140187     2.7367241   FALSE  
accel_arm_z           1.128000     4.0362858   FALSE  
magnet_arm_x          1.000000     6.8239731   FALSE  
magnet_arm_y          1.056818     4.4439914   FALSE  
magnet_arm_z          1.036364     6.4468454   FALSE  
roll_dumbbell         1.022388    84.2065029   FALSE  
pitch_dumbbell        2.277372    81.7449801   FALSE  
yaw_dumbbell          1.132231    83.4828254   FALSE  
total_accel_dumbbell  1.072634     0.2191418   FALSE  
gyros_dumbbell_x      1.003268     1.2282132   FALSE  
gyros_dumbbell_y      1.264957     1.4167771   FALSE  
gyros_dumbbell_z      1.060100     1.0498420   FALSE  
accel_dumbbell_x      1.018018     2.1659362   FALSE  
accel_dumbbell_y      1.053061     2.3748853   FALSE  
accel_dumbbell_z      1.133333     2.0894914   FALSE  
magnet_dumbbell_x     1.098266     5.7486495   FALSE  
magnet_dumbbell_y     1.197740     4.3012945   FALSE  
magnet_dumbbell_z     1.020833     3.4451126   FALSE  
roll_forearm         11.589286    11.0895933   FALSE  
pitch_forearm        65.983051    14.8557741   FALSE  
yaw_forearm          15.322835    10.1467740   FALSE  
total_accel_forearm   1.128928     0.3567424   FALSE  
gyros_forearm_x       1.059273     1.5187035   FALSE  
gyros_forearm_y       1.036554     3.7763735   FALSE  
gyros_forearm_z       1.122917     1.5645704   FALSE  
accel_forearm_x       1.126437     4.0464784   FALSE  
accel_forearm_y       1.059406     5.1116094   FALSE  
accel_forearm_z       1.006250     2.9558659   FALSE  
magnet_forearm_x      1.012346     7.7667924   FALSE  
magnet_forearm_y      1.246914     9.5403119   FALSE  
magnet_forearm_z      1.000000     8.5771073   FALSE  
classe                1.469581     0.0254816   FALSE  
```

### Model Selection and Fit Analysis
The training data set was partitioned into a training set (train) and testing set (test), with 70% in the training set and 30% in the testing set.

```{r, eval=FALSE}
set.seed(73821)
inTrain<-createDataPartition(training$classe, p=.7, list=F)
train <- training[inTrain,]
test <- training[-inTrain,]
```

The train data was used to build a Random Forest model with the number of trees limited to 50 to improve the run time. The Random Forrest model was chosen due to the non-linear relationship between the variables and the accuracy of these types of models.

```{r, eval=F}
modFit <- train(classe ~., data=train, method="rf", ntree=50, prox=TRUE)
```

The resulting model (modFit) provided an accuracy of 98.5% with an Kappa of .981. The model results were:

```{r, eval=F}
modFit
```

```
Random Forest  

11776 samples  
   52 predictor  
    5 classes: 'A', 'B', 'C', 'D', 'E'   
  
No pre-processing  
Resampling: Bootstrapped (25 reps)   
  
Summary of sample sizes: 11776, 11776, 11776, 11776, 11776, 11776, ...   
  
Resampling results across tuning parameters:  
  
  mtry  Accuracy  Kappa  Accuracy SD  Kappa SD  
   2    0.983     0.979  0.00199      0.00251   
  27    0.985     0.981  0.00252      0.00319   
  52    0.978     0.973  0.00359      0.00454   
  
Accuracy was used to select the optimal model using  the largest value.  
The final value used for the model was mtry = 27.   
```
  
The variable importance was then checked to determine if any variable could be removed from the model. Even though the importance begins to drop off after the first 20 variables as seen using the varImp function, it was decided to keep all variables. 
 
```{r, eval=FALSE}
varImp(modFit)
```

```
only 20 most important variables shown (out of 52)  
  
                     Overall  
roll_belt            100.000  
pitch_forearm         59.115  
yaw_belt              53.461  
magnet_dumbbell_z     47.532  
magnet_dumbbell_y     45.956  
pitch_belt            45.737  
roll_forearm          44.018  
accel_dumbbell_y      22.991  
roll_dumbbell         19.280  
accel_forearm_x       18.622  
magnet_belt_z         16.894  
magnet_dumbbell_x     15.624  
total_accel_dumbbell  14.300  
magnet_forearm_z      14.246  
magnet_belt_y         14.240  
accel_dumbbell_z      14.193  
accel_belt_z          13.410  
gyros_belt_z          13.036  
yaw_arm               10.787  
magnet_belt_x          9.835  
```
### Out of Sample Error Expectation
The train data, which was the 30% partition from the original Training, was then used to estimate the out of sample error of the model. The confusionMatix indicates a very good fit with an accuracy of 99% and a 95% confidence interval for the accuracy of [0.989, 0.993].

```{r, eval=FALSE}
pred_train <- predict(modFit, train)
confusionMatrix(pred_train, train$classe)
```

```
Confusion Matrix and Statistics  
  
          Reference  
Prediction    A    B    C    D    E  
         A 2231   12    0    0    0  
         B    1 1495    9    0    1  
         C    0   11 1356   27    0  
         D    0    0    3 1258    4  
         E    0    0    0    1 1437  
  
Overall Statistics  
                                          
               Accuracy : 0.9912          
                 95% CI : (0.9889, 0.9932)  
    No Information Rate : 0.2845            
    P-Value [Acc > NIR] : < 2.2e-16         
                                          
                  Kappa : 0.9889          
 Mcnemar's Test P-Value : NA              
  
Statistics by Class:  
  
                     Class: A Class: B Class: C Class: D Class: E  
Sensitivity            0.9996   0.9848   0.9912   0.9782   0.9965  
Specificity            0.9979   0.9983   0.9941   0.9989   0.9998  
Pos Pred Value         0.9947   0.9927   0.9727   0.9945   0.9993  
Neg Pred Value         0.9998   0.9964   0.9981   0.9957   0.9992  
Prevalence             0.2845   0.1935   0.1744   0.1639   0.1838  
Detection Rate         0.2843   0.1905   0.1728   0.1603   0.1832  
Detection Prevalence   0.2859   0.1919   0.1777   0.1612   0.1833  
Balanced Accuracy      0.9987   0.9916   0.9927   0.9886   0.9982  
```

### Predictions against Testing Data
Using the model defined above, the weight lifting correct/incorrectness was predicting for the 20 observations in the test set. The predicted results were

```{r, eval=FALSE}


testing_prediction <- predict(RF_model, test)
testing_prediction <- as.character(testing_prediction)
testing_prediction

pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(testing_prediction)
testing_prediction
```

```
[1] "B" "A" "B" "A" "A" "E" "D" "B" "A" "A" "B" "C" "B" "A" "E" "E" "A" "B" "B" "B"
```

When these values were submitted for project results, all 20 of the 20 were evaluated as correct.

### Conclusion
Given accelerometer data from a barbell lifting exercise being performed, it is possible to accurate determine if the exercise was performed correctly, and if not correct, in which way was it incorrect. It is expected that the same success could likely be expected from other weight lifting exercises, however it was not tested using this model. 

